---
title: "Getting Started with insperplot"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with insperplot}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.asp = 0.618,
  fig.align = "center",
  warning = FALSE,
  message = FALSE
)
```

## Introduction

**insperplot** extends ggplot2 with [Insper](https://www.insper.edu.br/en/)'s visual identity, providing custom themes, color palettes, and specialized plotting functions for academic and institutional use.

### Installation

```{r eval=FALSE}
# Install from GitHub
remotes::install_github("viniciusoike/insperplot")
```

```{r setup}
library(insperplot)
library(ggplot2)
library(dplyr)
```

### Font Setup (Optional but Recommended)

insperplot uses fonts based on Insper's official template: **Georgia** (primary for titles), **Inter** and **Arial** (for body text). **EB Garamond** and **Playfair Display** are used as fallbacks for titles. While optional, installing these fonts significantly enhances plot aesthetics.

Note: "GT Ultra Fine Bold" is one of the official Insper fonts, but it is not freely available and is not included in the package.

```{r eval=FALSE}
# Check font status
setup_insper_fonts()

# Download missing fonts automatically
import_insper_fonts()
```

**Manual installation:**

1. Visit [Google Fonts](https://fonts.google.com)
2. Download "Inter", "EB Garamond", and "Playfair Display"
3. Install on your system
4. Restart R/RStudio

Georgia is typically pre-installed on most systems. If fonts are unavailable, plots will use system defaults.

## Basic Usage: theme_insper()

The foundation of insperplot is `theme_insper()`, which applies Insper's visual identity to any ggplot2 plot. Simply add it to any ggplot2 chart to get a clean, branded appearance with Insper's typography and colors.

```{r theme-basic}
ggplot(mtcars, aes(mpg, wt)) +
  geom_point() +
  labs(
    title = "Basic Theme",
    subtitle = "Using theme_insper()",
    caption = "Source: mtcars"
  ) +
  theme_insper()
```

### Customizing the Theme

`theme_insper()` supports several arguments for quick customization:

- `font_title`: Font for titles. Default is `"Georgia"`.
- `font_text`: Font for body text. Default is `"Inter"`.
- `grid`: Grid lines. Default is `TRUE`.
- `border`: Border style. Options are `"none"` (default), `"half"`, or `"closed"`.
- `align`: Title, subtitle, and caption alignment. Default is `"panel"`.

The following examples use `macro_series`, a dataset bundled with insperplot. It contains monthly Brazilian macroeconomic indicators — including inflation (IPCA), industrial production (IPI), and economic activity (IBC-Br) — sourced from the [Brazilian Central Bank](https://www3.bcb.gov.br/sgspub/). See `?macro_series` for details.

Here we plot the IBC-Br index (a monthly GDP proxy) with a `"half"` border and a sans-serif title font:

```{r theme-border-half}
ggplot(macro_series, aes(date, ibcbr_dessaz)) +
  geom_line(lwd = 0.7) +
  labs(
    title = "Economic Activity Index",
    subtitle = "IBC-Br, seasonally adjusted (base 2002 = 100)",
    x = NULL,
    y = "IBC-Br",
    caption = "Source: Banco Central do Brasil"
  ) +
  theme_insper(font_title = "Inter", border = "half")
```

### Fine-Tuning with theme_sub_*()

For more granular control, ggplot2 4.0+ provides `theme_sub_*()` helper functions that target specific theme components. Add them after `theme_insper()` to override individual elements without resetting the whole theme.

In the example below, we add axis ticks and lines to the x-axis with `theme_sub_axis_x()`, and change the panel background with `theme_sub_panel()`:

```{r theme-sub-override}
ggplot(macro_series, aes(date, ibcbr_dessaz)) +
  geom_line(lwd = 0.7) +
  labs(
    title = "Economic Activity Index",
    x = NULL,
    y = "IBC-Br",
    caption = "Source: Banco Central do Brasil"
  ) +
  theme_insper() +
  theme_sub_axis_x(
    ticks = element_line(color = "gray20", linewidth = 0.25),
    line = element_line(color = "gray20", linewidth = 0.5)
  ) +
  theme_sub_panel(
    background = element_rect(fill = "#f3f3f3", color = NA)
  )
```

## Working with Colors

The colors used in insperplot are based on Insper's [brandkit](https://github.com/Insper/Cores-Marca-Insper). Use `insper_palette()` to preview a specific palette and `show_insper_palettes()` to see all available palettes at a glance:

```{r palettes-overview, fig.height=4}
insper_palette("reds")

show_insper_palettes()
```

### Using Colors in ggplot2

The package provides `scale_color_insper_*()` and `scale_fill_insper_*()` functions for integrating Insper colors into ggplot2 plots. Use the `_d` variants for discrete (categorical) data and `_c` for continuous (numeric) data.

#### Discrete Scales

For categorical variables, use `scale_color_insper_d()` or `scale_fill_insper_d()`.

The examples below use `fossil_fuel`, a dataset bundled with insperplot. It tracks global primary energy consumption (in TWh) by fuel type (coal, oil, and gas) from 1800 to the present, sourced from [Our World in Data](https://ourworldindata.org/grapher/global-fossil-fuel-consumption). See `?fossil_fuel` for details.

First, a stacked area chart of consumption since 1960 using `scale_fill_insper_d()`:

```{r scale-discrete-fill}
recent_fuel <- subset(fossil_fuel, year >= 1960)

ggplot(recent_fuel, aes(year, consumption, fill = fuel)) +
  geom_area() +
  scale_fill_insper_d(palette = "bright") +
  scale_x_continuous(expand = expansion(mult = c(0))) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.05))) +
  theme_insper(border = "half")
```

The same data shown as a line chart with `scale_color_insper_d()`:

```{r scale-discrete-color}
ggplot(fossil_fuel, aes(year, consumption, color = fuel)) +
  geom_line() +
  scale_color_insper_d(palette = "bright") +
  theme_insper() +
  labs(
    title = "Global Fossil Fuel Consumption",
    x = NULL,
    y = "Consumption (TWh)",
    color = "Fuel",
    caption = "Source: Our World in Data (CC BY 4.0)"
  )
```

#### Continuous Scales

For numeric variables, use `scale_color_insper_c()` or `scale_fill_insper_c()`. These create smooth color gradients from any Insper palette.

The heatmap below shows 12-month rolling inflation (IPCA) across years and months using the `"reds"` palette. We compute the rolling product from the `macro_series` dataset using `lubridate` (for date parts) and `RcppRoll` (for the rolling window):

```{r scale-continuous-fill}
#| fig-width: 8
#| fig-height: 9
macro_calendar <- macro_series |>
  mutate(
    year = lubridate::year(date),
    month = lubridate::month(date),
    ipca12m = RcppRoll::roll_prodr(1 + ipca / 100, n = 12) - 1,
    ipca12m = ipca12m * 100
  ) |>
  filter(between(year, 2004, 2024))

ggplot(macro_calendar, aes(month, year, fill = ipca12m)) +
  geom_tile(color = "white", lwd = 0.1) +
  scale_x_continuous(
    breaks = 1:12,
    labels = month.abb,
    position = "top",
    expand = expansion(c(0))
  ) +
  scale_y_continuous(breaks = 2004:2024, expand = expansion(c(0))) +
  scale_fill_insper_c(
    name = "IPCA (YoY, %)",
    palette = "reds"
  ) +
  labs(x = NULL, y = NULL) +
  theme_insper() +
  theme_sub_legend(
    title = element_text(family = "Georgia", size = 12),
    title.position = "top",
    key.width = unit(4, "lines"),
    position = "top"
  )
```

## Convenience Plot Functions

Beyond the theme and scales, insperplot provides high-level plotting functions that combine common geom + scale + theme patterns into a single call. Each function returns a standard ggplot object, so you can still add layers, labs, or further customization with `+`.

### Time Series

`insper_timeseries()` creates line plots optimized for temporal data. It handles `Date` and `POSIXct` axes automatically and supports multiple series via the `color` argument.

```{r plot-timeseries}
recent_macro <- subset(macro_series, date >= as.Date("2015-01-01"))

insper_timeseries(recent_macro, x = date, y = ipca) +
  labs(
    title = "Brazilian Inflation (IPCA)",
    subtitle = "Monthly rate, 2015–present",
    x = NULL,
    y = "IPCA (%)",
    caption = "Source: Banco Central do Brasil"
  )
```

## Saving Plots

For best results when saving plots, use `ggsave()` with the [ragg](https://ragg.r-lib.org/) device to ensure proper font rendering:

```{r save-ggsave, eval=FALSE}
p <- ggplot(mtcars, aes(wt, mpg)) +
  geom_point() +
  theme_insper()

ggsave(
  "my_plot.png",
  p,
  width = 8,
  height = 5,
  dpi = 300,
  device = ragg::agg_png
)
```

Alternatively, `save_insper_plot()` wraps `ggsave()` with sensible defaults for Insper-branded output: dimensions in centimeters, a golden-ratio aspect ratio, and automatic ragg device selection for PNG files when the package is installed.

```{r save-insper, eval=FALSE}
save_insper_plot(p, "my_plot.png")

# Custom dimensions (in cm by default)
save_insper_plot(p, "my_plot.png", width = 16, height = 10)
```

## Available Functions

In addition to `insper_timeseries()` shown above, insperplot includes convenience functions for other common chart types:

| Function | Chart type |
|---|---|
| `insper_timeseries()` | Line plots for temporal data |
| `insper_barplot()` | Vertical and horizontal bar charts |
| `insper_scatterplot()` | Scatter plots |
| `insper_area()` | Area charts |
| `insper_boxplot()` | Box plots |
| `insper_violin()` | Violin plots |
| `insper_histogram()` | Histograms |
| `insper_density()` | Density plots |
| `insper_heatmap()` | Heatmaps |

All functions return ggplot objects, apply `theme_insper()` by default, and accept a `palette` argument for Insper color palettes. See their help pages (e.g. `?insper_barplot`) for full documentation.

## Learn More

- [Package website](https://viniciusoike.github.io/insperplot/) — full function reference and additional articles
- [GitHub repository](https://github.com/viniciusoike/insperplot) — source code, issues, and contributing
- [Insper brand colors](https://github.com/Insper/Cores-Marca-Insper) — the color system this package is built on
